---
title: "Voting Difficulty as a Rift between Democrats and Republicans"
author: "Samantha Williams, Varune Maharaj, Wilford Bradford, Dmitri Zadvornov"
date: "3/02/2022"
output:
  pdf_document:
    toc: yes
  bookdown::pdf_document2:
    toc: yes
    number_sections: yes
    fig_caption: yes
  html_document:
    toc: yes
    number_sections: yes
    fig_caption: yes
    df_print: paged
---
\newpage
\setcounter{page}{1}

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# load required libraries
library(dplyr)
library(ggplot2)
library(tidyverse)
library(Hmisc) # descriptive
library(psych, exclude = c('describe','alpha','%+%')) # descriptive by groups
library(coin) # for effect size calculation
library(rstatix, exclude = 'filter') # effect size
library(knitr)

theme_set(theme_bw())
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo=FALSE, message=FALSE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, show_col_types = FALSE}
# load data set
df_init <- read_csv("anes_timeseries_2020_csv_20220210.csv")

```
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# conduct recoding, new variable calculation, wrangling
#
# new column names after mapping of given column names
eda <- c('case_id',
         'party_id', 
         'voted', 
         'voting_difficulty',
         'vproblem_registration',
         'vproblem_id',
         'vproblem_abs_ballot',
         'vproblem_conf_ballot',
         'vproblem_transport',
         'vproblem_wait',
         'vproblem_work',
         'vproblem_weather',
         'vproblem_mail_ballot',
         'vproblem_other',
         'nvproblem_main',
         'nvproblem_other',
         'vpractical_difficulty',
         'nvcombined_difficulty'
)

df_min <- df_init %>% rename(
  case_id = 'V200001',
  #pre_party_registration = 'V201018', #binary based on interest
  #post_party_registration = 'V202064',
  party_id_sum = 'V201231x', # 0=no answer, 1-3=Democrat, 4=Independent, 5-7=Republican
  voted = 'V202109x', #binary 0=no vote, 1=vote, -2=not reported
  voting_difficulty = 'V202119', #ordinal less than 0=No info, 1=Not difficult, 2=Little difficult, 3=Moderately difficult, 4=Very difficult, 5=Extremely Difficult
  vproblem_registration = 'V202120a', #0=Not mentioned, 1=Mentioned, REGISTRATION PROBLEM
  vproblem_id = 'V202120b', #CONCERN ABOUT IDENTIFICATION CARD
  vproblem_abs_ballot = 'V202120c', #DIFFICULTY OBTAINING ABSENTEE BALLOT
  vproblem_conf_ballot = 'V202120d', #CONFUSION ABOUT BALLOT OR MACHINE
  vproblem_transport = 'V202120e', #DIFFICULTY GETTING TO POLLING PLACE
  vproblem_wait = 'V202120f', #LONG WAIT TIMES
  vproblem_work = 'V202120g', #WORK SCHEDULE
  vproblem_weather = 'V202120h', #BAD WEATHER
  vproblem_mail_ballot = 'V202120i', #ISSUE MAILING BALLOT
  vproblem_other = 'V202120j', #OTHER PROBLEM # vproblem_none = 'V202120k', #NONE AT ALL
  nvproblem_main = 'V202123', #MAIN REASON R DID NOT VOTE, 1=I forgot, 2=I’m not interested, 3=Too busy, 4=Did not like the candidates, 5=I am not registered, 6=I did not have the correct form of identification, 7=Out of town, 8=Sick or disabled, 9=Transportation, 10=Bad weather, 11=The line at the polls was too long, 12=I was not allowed to vote at the polls, even though I tried 13. I requested but did not receive an absentee ballot, 14=I did not know where to vote, 15=I did not feel that I knew enough about the choices, 16=Other {SPECIFY}
  nvproblem_other = 'V202124', #OTHER REASON R DID NOT VOTE, 1=I forgot, 2=I’m not interested, 3=Too busy, 4=Did not like the candidates, 5=I am not registered, 6=I did not have the correct form of identification, 7=Out of town, 8=Sick or disabled, 9=Transportation, 10=Bad weather, 11=The line at the polls was too long, 12=I was not allowed to vote at the polls, even though I tried 13. I requested but did not receive an absentee ballot, 14=I did not know where to vote, 15=I did not feel that I knew enough about the choices, 16=Other {SPECIFY}, 17=There were no other reasons {VOL}
) %>%
# Mutate/recode data for statistical tests
  mutate(
     party_id = # Recode Party ID Summary to text labels
      case_when(
        party_id_sum == -9
        | party_id_sum == -8
        ~ 'Unknown',
        party_id_sum == 4
        ~ 'Independent',
        party_id_sum == 1
        | party_id_sum == 2
        | party_id_sum == 3
        ~ 'Democrat',
        party_id_sum == 5
        | party_id_sum == 6
        | party_id_sum == 7
        ~ 'Republican'),
    # calculate vpractical_difficulty as amount of reported difficulties
    # this only applies for respondents with status of voted (voted == 1)
    vpractical_difficulty = (voted == 1) * ((vproblem_registration * 
                                               (vproblem_registration >= 0)
       + vproblem_id * (vproblem_id >= 0)
       + vproblem_abs_ballot * (vproblem_abs_ballot >= 0)
       + vproblem_conf_ballot * (vproblem_conf_ballot >= 0)
       + vproblem_transport * (vproblem_transport >= 0)
       + vproblem_wait * (vproblem_wait >= 0)
       + vproblem_work * (vproblem_work >= 0)
       + vproblem_weather * (vproblem_weather >= 0)
       + vproblem_mail_ballot * (vproblem_mail_ballot >= 0))),
    # calculate nvcombined_difficulty, which is combination of responses by
    # respondents for main and other reasons that caused them to not vote 
    # (vote == 0). The values for each provided reason are counted as follows,
    # given the subjectively evaluated factor of intent - 
    # The definition of difficulty applied here is
    # - respondent made all reasonable efforts to participate in the voting, but
    # had a difficulty in doing so. Essentially, skipping options that indicate
    # no intent to vote.
    #
    # 1. I forgot - 0 - no intent
    # 2. I’m not interested - 0 - no intent
    # 3. Too busy - 0 - no intent
    # 4. Did not like the candidates - 0 - no intent
    # 5. I am not registered - 1
    # 6. I did not have the correct form of identification - 1
    # 7. Out of town - 0 - no intent (under condition of absentee ballot avail)
    # 8. Sick or disabled - 1
    # 9. Transportation - 1
    # 10. Bad weather - 1
    # 11. The line at the polls was too long - 1
    # 12. I was not allowed to vote at the polls, even though I tried - 1
    # 13. I requested but did not receive an absentee ballot - 1
    # 14. I did not know where to vote - 1
    # 15. I did not feel that I knew enough about the choices - 1
    # 16. Other {SPECIFY} - 0 - intent unknown <=> no intent
    nvcombined_difficulty = #main/other - 0/0 -> 0, 0/1 -> 1, 1/0 -> 2, 1/1 -> 3
      (voted == 0) * 2 *
      case_when(
        nvproblem_main >= 5
        & nvproblem_main <= 15
        & nvproblem_main != 7
        ~ 1,
        TRUE ~ 0) + # code "not at all" or "reject to respond" as 0
     case_when(
        nvproblem_other >= 5
        & nvproblem_other <= 15
        & nvproblem_other != 7
        ~ 1,
        TRUE ~ 0) # code "not at all" or "reject to respond" as 0
  ) %>%
select(all_of(eda)) %>% 
  # Next, the data is filtered to remove the rows with non-applicable data - 
  # the rows, where interview ended prematurely based on indicators
  # of key variables:
  # 1. voted, but did not provide voting_difficulty or ended interview before
  #    providing responses to all the difficulty prompts
  # 2. did not vote and ended interview before responding to difficulty prompts.
  # 3. ended interview before voting response was provided
  filter(!((voted == 1
           & (voting_difficulty < 0
             | (vproblem_registration <0 & vproblem_registration != -9)
             | (vproblem_id <0 & vproblem_id != -9)
             | (vproblem_abs_ballot <0 & vproblem_abs_ballot != -9)
             | (vproblem_conf_ballot <0 & vproblem_conf_ballot != -9)
             | (vproblem_transport <0 & vproblem_transport != -9)
             | (vproblem_wait <0 & vproblem_wait != -9)
             | (vproblem_work <0 & vproblem_work != -9)
             | (vproblem_weather <0 & vproblem_weather != -9)
             | (vproblem_mail_ballot <0 & vproblem_mail_ballot != -9))
           )
           | (voted == 0
              & ((nvproblem_main < 0 
                  & nvproblem_main != -8
                  & nvproblem_main != -9)
                 | (nvproblem_other < 0
                    & nvproblem_other != -8
                    & nvproblem_other != -9))
           )
           | (voted < 0)))
               
```
# Importance and Context
A cornerstone of any democracy is voting rights. A person is eligible to vote if 
they are 18 years old on or before Election Day, a U.S. citizen, and registered 
to vote, with some states having additional requirements for voter eligibility. 
However, there is a long history of tactics that have been used to make voting 
difficult for various groups. Literacy tests, poll taxes, voter 
disenfranchisement laws, and fraud are all examples of the tactics used by the 
two major parties, Democratic and Republican.^[ACLU. “Block the Vote: How 
Politicians are Trying to Block Voters from the Ballot Box” (2021)] Generally 
speaking, a voter who identifies as a Republican has conservative social values, 
prefers a smaller government, and wants to reduce taxes to stimulate the 
economy, while a voter who identifies as a Democrat tends to have liberal 
social values, believes that government should be the safety net for all and 
prefers a progressive tax structure. 
During the 2020 US Presidential Election, there were claims made by both 
parties of increased measures to make it more difficult to cast a vote. Efforts 
to provide safe voting options for all due to the COVID-19 global pandemic, may 
have been a contributing factor to additional concerns about voting 
difficulty.^[Levinson-King, R. “US election 2020: Why it can be hard to vote in 
the US” (2020)] This led to voters questioning their voting experience and 
wondering if the election process was manipulated for political gain. To gain 
further insight, this analysis aims to address the following research question:  
\begin{quote}
  \textit{Did Democratic voters or Republican voters experience more difficulty 
  voting in the 2020 election?}
\end{quote}
The answer to this question can be pivotal in understanding the voter 
experience during a global pandemic and if claims that a particular party did 
have a more difficult voting experience compared to the other. Additionally, it 
would provide more context for accusations of voter suppression and further 
clarify if a particular party gained a political advantage from these tactics.

# Data and Methodology
Our analyses will leverage the data from the 2020 American National Election 
Studies (ANES), which is a continuation of the series of election studies 
conducted since 1948 to support analysis of public opinion and voting behavior 
in U.S. presidential elections. The 2020 study features re-interviews with 2016 
ANES respondents, a freshly drawn cross-sectional sample, and post-election 
surveys with respondents from the General Social Survey (GSS). The study has a 
total of 8,280 pre-election interviews and 7,449 post-election re-interviews 
satisfying the criteria of IID. 
 
As previously discussed, there are two major parties in the US; other parties 
make up a very small percentage and are not considered in this analysis. The 
respondents are asked for the pre and post-vote party affiliation, as well as 
questions around the party preference for various offices. Based on the study 
conducted by Petrocik, voters that initially identified 
themselves as independent did so to highlight their independent mindedness as 
opposed to their party affiliation.^[Petrocik “Measuring party support: Learners
are not independents” (2009)] Further probing with follow-up questions 
in most cases identified a particular leaning of the respondent toward specific 
affiliation with varying degree of preference. For our purposes we ignore this 
and define voter’s affiliation as leaning within any degree of preference 
toward a specific party. This made the variable V201231x in the ANES data, 
named party_id in our analysis, as the best available method to summarize the 
respondent’s lean. It is an unpaired ordinal scale that relates to party 
affiliation using a range of 1 (Democrat) to 7 (Republican) with 4 as the 
neutral Independent Voter. Table 1 gives a summary of participant affiliation, 
and shows no significant difference between the two major parties.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
kable(
  t(summary(as.factor(df_min$party_id))),
  digits = 2,
  caption = 'Party ID', 
  booktabs = TRUE
)
``` 
We identified voting difficulty by analyzing the V202119 (aka voting_difficulty)
field that asked the respondent how difficult it was to cast their vote based on
an ordinal scale. The survey further expanded the voter difficulty question by 
asking binary follow up questions (V202120 parts a-j aka practical_difficulty) 
to identify where the voter had difficulty with: registering to vote, obtaining 
a voter ID card, obtaining an absentee ballot, confusion about the ballot or 
machine, work schedule conflicts, bad weather, getting to polling location, 
issues mailing their ballot, long wait times to vote or other difficulties 
experienced.
```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(
  data.frame (
    variable  = c(
      "V202120a", 
      "V202120b",
      "V202120c",
      "V202120d",
      "V202120e",
      "V202120f",
      "V202120g",
      "V202120h",
      "V202120i",
      "V202120j"
    ),
    description = c(
      "Difficulty Registering to Vote", 
      "Obtaining an Acceptable Voter ID Card",
      "Obtaining an Absentee Ballot",
      "Confusion about the Ballot or Machine",
      "Difficulty Getting to Polling Locations",
      "Long Wait Times",
      "Work Schedule",
      "Bad Weather",
      "Issue Mailing Ballot",
      "Other Problem"
    )
  ),
  digits = 2,
  caption = 'Practical Difficulty', 
  booktabs = TRUE
)
```
To complete our analysis we used the variable V202109 that is a binary question 
that informs if the respondent voted or not. We found that 86.13% voted while 
13.87% did not. Since the group of respondents who did not vote did not answer 
the voting_difficulty question, we wondered if anyone in this subsection 
experienced such difficulty that they could not or were deterred from casting 
their vote. To find out we looked at the variables V202123 and V202124 that 
asked the main reason the respondent did not vote. Given the subjectively 
evaluated factor of intent, the definition of difficulty applied to our analysis
is that respondents made all reasonable efforts to participate in the voting but
had difficulty in doing so, we have skipped response options that allude to no 
intent to vote.

We sought to determine whether those who responded to the survey did encounter 
difficulty voting and if those who intended to vote also experienced some degree
of difficulty based on party affiliation. The One-Sided Wilcoxon Rank-Sum Test 
was determined to be the best choice based on the assumptions of ordinal, 
unpaired data and shapes as supported by Figure 1.
```{r fig.show="hold", out.width="50%", echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Voting Difficulty by Party ID Distribution Shapes"}
#Are they identically distributed? 
hist(as.numeric(df_min$voting_difficulty[df_min$party_id == 'Democrat']), 
     main='Democrat Voters', xlab="Perceived Difficulty")
hist(as.numeric(df_min$voting_difficulty[df_min$party_id == 'Republican']), 
     main='Republican Voters', xlab="Perceived Difficulty")

```
\newpage
We defined the following hypotheses for the tests:

**Null Hypothesis:** _For voters in the 2020 election who identified as Democrats there was no reported difference in voting difficulty compared to Republican voters._ 
 
**Alternative Hypothesis:** _For voters in the 2020 election who identified as Democrats there was a reported difference in voting difficulty compared to Republican voters._

In order to try and capture the effect of voting difficulty between parties we conducted the following tests: The first test looked at non-voters and non-voted combined difficulty (aka nvcombined_difficulty), derived through binary calculation as follows: $$2\cdot I_{main} + I_{other}$$ The second test was based directly on self-reported (i.e. perceived) voting difficulty (aka voting_difficulty). The third test utilized practical difficulty (aka practical_difficulty) that we defined as the overall count of difficulty indicators the respondent identified in the survey.

```{r, echo=FALSE, warning=FALSE, message=FALSE, results=FALSE}
# filter down data to only Democrat or Republican voters that did not vote
# conduct the test on nvcombined_difficulty variable, using party id as a 
# factor.
# The test we run is one-sided under alternative of having greater difficulty
# for Democrats to vote.
# H0: Democrats had less or the same difficulty voting as Republicans
# Alpha = 0.05
df_nv <- df_min %>%
  filter((party_id == "Democrat" | party_id == "Republican") & voted == 0)

wilcox.test(as.numeric(nvcombined_difficulty) ~ as.factor(party_id), data=df_nv, alternative = 'g')

```

```{r, echo=FALSE, warning=FALSE, message=FALSE, results=FALSE}
# filter down data to only Democrat or Republican voters that did vote
# conduct the test on voting_difficulty variable (perceived), using party id 
# as a factor.
# The test we run is one-sided under alternative of having greater difficulty
# for Democrats to vote.
# H0: Democrats had less or the same difficulty voting as Republicans
# Alpha = 0.05
#
df_v <- df_min %>%
  filter((party_id == "Democrat" | party_id == "Republican") & voted == 1)

wilcox.test(as.numeric(voting_difficulty) ~ as.factor(party_id), data=df_v, 
            alternative = 'g')

```
```{r, echo=FALSE, warning=FALSE, message=FALSE, results=FALSE}
# effect size
wilcox_effsize(voting_difficulty ~ party_id, data=df_v, 
            alternative = 'g')
```

``` {r, echo=FALSE, warning=FALSE, message=FALSE, results=FALSE}
# Percentages for each voter identity group w/ little difficulty or above 
sum(df_v$voting_difficulty[df_v$party_id == 'Democrat'] > 1)/length(df_v$voting_difficulty[df_v$party_id == 'Democrat']) - 
sum(df_v$voting_difficulty[df_v$party_id == 'Republican'] > 1)/length(df_v$voting_difficulty[df_v$party_id == 'Republican'])
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, results=FALSE}
# filter down data to only Democrat or Republican voters that did vote
# conduct the test on vpractical_difficulty variable (reported number of 
# difficulties), using party id as a factor.
# The test we run is one-sided under alternative of having greater difficulty
# for Democrats to vote.
# H0: Democrats had less or the same difficulty voting as Republicans
# Alpha = 0.05
#
df_v <- df_min %>%
  filter((party_id == "Democrat" | party_id == "Republican") & voted == 1)

wilcox.test(as.numeric(vpractical_difficulty) ~ as.factor(party_id), data=df_v, 
            alternative = 'g')

```

```{r checking counts of available data to work with, echo=FALSE, warning=FALSE, message=FALSE, results=FALSE}
sum(df_min$voted & df_min$party_id == 'Democrat')
sum(df_min$voted & df_min$party_id == 'Republican')
sum(df_min$voted & df_min$party_id == 'Unknown')
```

```{r counting those who responded voting difficulty and didnt vote, echo=FALSE, warning=FALSE, message=FALSE, results=FALSE}
sum(df_min$voted == 1 & df_min$voting_difficulty > 0, na.rm=TRUE)
# did not vote and stated that they had some kind of difficulty voting
sum(df_min$voted == 0 & df_min$voting_difficulty > 0, na.rm=TRUE)

#what are the sample sizes between the two parties? 
sum(df_min$party_id == 'Democrat' & df_min$voting_difficulty > 0, na.rm=TRUE)
sum(df_min$party_id == 'Republican' & df_min$voting_difficulty > 0, na.rm=TRUE)
```

```{r filering information where a result for voting difficulty was provided, echo=FALSE, warning=FALSE, message=FALSE, results=FALSE}
# how much data will be working with? 
filter(df_min, party_id == 'Republican' | party_id == 'Democrat')
```

```{r fig.show="hold", out.width="50%", echo=FALSE, warning=FALSE, message=FALSE, fig.show='hide', results=FALSE}
#Are they identically distributed? 
hist(as.numeric(df_min$nvcombined_difficulty[df_min$party_id == 'Democrat']), 
     main='Democrat Non-Voters', xlab="Difficulty")
hist(as.numeric(df_min$nvcombined_difficulty[df_min$party_id == 'Republican']), 
     main='Republican Non-Voters', xlab="Difficulty")
```

```{r fig.show="hold", out.width="50%", echo=FALSE, warning=FALSE, message=FALSE, fig.show='hide', results=FALSE}
#Are they identically distributed? 
hist(as.numeric(df_min$vpractical_difficulty[df_min$party_id == 'Democrat']), 
     main='Democrat Voters', xlab="Practical Difficulty")
hist(as.numeric(df_min$vpractical_difficulty[df_min$party_id == 'Republican']), 
     main='Republican Voters', xlab="Practical Difficulty")
```
# Results
Based on the defined hypothesis we ran three tests with the following results: 

**Non-voted combined difficulty indicator test results:** Based on the p-value 
of 0.3242 we have **failed to reject** the null toward the alternative 
hypothesis, that it is more difficult to vote for Democrat voters (vs. 
Republican) for the participants under the unweighted sub-population that did 
not vote with significance level $\alpha$ of 5%.
 
**Voted (perceived, self-reported) difficulty indicator test results:** We 
obtained a statistically significant test result based on the p-value of 
0.00177. As such, we have **rejected** the null toward the alternative 
hypothesis, that it is more difficult to vote for Democrat voters (vs. 
Republican) for the participants under the unweighted sub-population that did 
vote with significance level $\alpha$ of 5%. Further, the effect size based on 
normal approximation is 3.8% which is small in magnitude. This is also supported
by the fact that there is only a 2.5% difference between fraction of Democrats 
and fraction of Republicans that experienced difficulty.

**Voted practical difficulty indicator test results:** Based on the p-value of 
0.23, though borderline, we have **failed to reject** the null toward the 
alternative hypothesis that it is more difficult to vote for Democrat voters 
(vs. Republican) for the participants under the unweighted sub-population that 
did vote with significance level $\alpha$ of 5%.

# Discussion
The analysis performed indicated that there was a statistically significant 
difference in difficulty voting between respondents who identified as Democrat 
versus Republicans. However, we could not find any practical significance for 
the discovered effect. 

These results are a strong indicator that voters had equitable access to cast 
their vote despite the challenges associated with unconventional voting 
methodologies incorporated due to the unique circumstances at the time. Not 
enough evidence was found to support the narrative that any party gained a 
political advantage in the election.


However, it is understood that there is an underlying complexity of difficulty 
experienced. It may be composed of cognitive, physical, or psychological 
factors. To improve research method, an instrument similar to the academic 
persistence scale should be utilized in the future to properly capture voting 
difficulty experienced by the respondents and improve the power of the test and 
potentially lead to stronger measure of effect.^[Thalib, Tarmizi & Hanafi, Sekar
& Asyraf, M Fahmi & Irbah, Shidqi & J.S, Eduardus. (2018). The Academic 
Persistence Scale.]



